---
title: "Introduction to the ess package"
author: "Jorge Cimentada"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ess for R and Stata users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Introduction

Using the `ess` package is fairly easy. There are are two main families of functions: `ess_*` and `show_*`. They each complement each other and allow the user to almost never have to go to the European Social Survey (ESS) website. The only scenario where you need to enter the ESS website is to validate your email. If you haven't registered, create an account at http://www.europeansocialsurvey.org/user/new

Once you register visit your email account to validate the account and you're ready to access the data.

Given that some `ess` functions require your email address, this vignette will use a fake email but everything should work accordingly if you registered with the ESS.

## Downloading country specific rounds

To install and load development version of the package use:
```{r}
install.packages("devtools")
devtools::install_github("cimentadaj/ess")

library(ess)
```

to install the stable version from CRAN use:

```{r}
install.packages("ess")
```

```{r, echo = FALSE, eval = TRUE}
library(ess)
```

Let's suppose you don't know which countries or rounds are available for the ESS. Then the `show_*` family of functions is your friend.

To find out which countries have participated you can use `show_countries()`

```{r, eval = TRUE}
show_countries()
```

This function actually looks up the countries in the ESS website. If new countries enter, this will automatically grab those countries as well. Let's checkout Turkey. How many rounds has Turkey participated in? We can use `show_country_rounds()`

```{r, eval = TRUE}
tk_rnds <- show_country_rounds("Turkey")
tk_rnds
```
Note that country names are case sensitive. Use the exact name printed out by `show_countries()`

Using this information, we can download those specific rounds easily with `ess_country`.

```{r}
turkey <-
  ess_country(
    country = "Turkey",
    rounds = c(2, 4),
    your_email = "your_email@random.com"
    )
```

`turkey` will now be a list of `length(rounds)` containing a data frame for each round. If you only specified one round, then all `ess_*` functions return a data frame. ess_country` is useful for when you want to download specific rounds, but not all. To download all rounds for a country automatically you can use `ess_all_cntrounds`.

```{r}
ess_all_cntrounds("Turkey", "your_email@random.com")
```

The `ess_*` family is  concerned with downloading the data and thus always returns a list containing data frames. Conversely, the `show_*` family grabs information from the ESS website and always returns vectors.

## Download complete rounds

Similarly, we can use other functions to download rounds. To see which rounds are currently available, use `show_rounds`.

```{r, eval = TRUE}
show_rounds()
```

Similar to `show_countries`, `show_rounds` interactively looks up rounds in the ESS website, so any future rounds will automatically be included.

To download all available rounds, use `ess_all_rounds`

```{r}
all_rounds <- ess_all_rounds("your_email@random.com")
```

Alternatively, use `ess_rounds` for selected ones.

```{r}
selected_rounds <- ess_rounds(c(1, 3, 6), "your_email@random.com")
```

## Downloading data for Stata, SPSS and SAS users

All `ess_*` functions allow the user to save the datasets in a specified folder in `'stata'`, `'spss'` or `'sas'` formats.

For example, to save round two from Turkey in a folder called `./my_folder`, we use:

```{r}
ess_country("Turkey", 2,
            "your_email@random.com",
            only_download = TRUE,
            output_dir = "./myfolder/")
```

By default it saves the data as `'stata'` files. Alternatively you can use `'spss'` or `'sas'`.

```{r}
ess_country("Turkey", 2,
            "your_email@random.com",
            only_download = TRUE,
            output_dir = "./myfolder/",
            format = 'sas')
```

This will save the data to `./myfolder/ESS_Turkey` and inside that folder there will be the `ESS2` folder that contains the data. If you specify `only_download = TRUE`, but don't specify a directory, an error will be raised. All of this applies exactly the same way for `ess_rounds`. If `only_download = TRUE` the format argument will be ignored, as it's only useful for downloading the data.

## Correcting for missing values

Whenever you download the ESS data, it comes together with a script that recodes the values 6 = 'Not applicable', 7 = 'Refusal', 8 = 'Don't know', 9 = 'No answer' and 9 = 'Not available' as missings. However, that is the case for variables that have a scaling of 1-5. For variables which have a scaling from 1-10 the corresponding missings are 66, 77, and so on. At first glance new users might not know this and start calculating statistics with these variables such as

```{r}
sp <- ess_country("Spain", 1, your_email)
mean(sp$tvtot)
# 4.622406
```

But that vector numbers such as `66`, `77`, that shouldn't be there. `recode_missings()` removes the corresponding missings for numeric variables as well as for character variables.

```{r}
new_coding <- recode_missings(sp)
mean(new_coding$tvtot)
# 4.527504
```

It also gives you the option of recoding only specific categories. For example

```{r}
other_newcoding <- recode_missings(sp, c("Don't know", "Refusal"))
table(other_newcoding$tvpol)
#  0   1   2   3   4   5   6   7  66 
# 167 460 610 252  95  36  26  31  45 
```
Still has missing values but recoded the ones you were interested in. I strongly suggest the user not to recode these categories as missing without looking at the data as there might be substantial diferences between people who didn't answer and answered questions. But if you're decided to do so, use `recode_missings` to recode everything and the `recode_*_missings` functions for numeric and character recodings separately. See the documentation of `recode_missings` for more information.

Currently the user can either download the data or use it in `R`. For example, you can't download the data into the specified folder and also get the list with the rounds.