---
title: "Introduction to the essurvey package"
author: "Jorge Cimentada"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using essurvey for R and Stata users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Introduction

Using the `essurvey` package is fairly easy. There are are two main families of functions: `import_*` and `show_*`. They each complement each other and allow the user to almost never have to go to the European Social Survey (ESS) website. The only scenario where you need to enter the ESS website is to validate your email. If you haven't registered, create an account at http://www.europeansocialsurvey.org/user/new. For those unfamiliar with the ESS, this vignette uses the term rounds, here a synonym of waves to denote the same survey in different time points.

Once you register visit your email account to validate the account and you're ready to access the data.

Given that some `essurvey` functions require your email address, this vignette will use a fake email but everything should work accordingly if you registered with the ESS.

## Downloading country specific rounds

To install and load development version of the package use:
```{r}
# install.packages("devtools")
devtools::install_github("cimentadaj/essurvey")
```

<!-- to install the stable version from CRAN use: -->

```{r, echo = FALSE}
install.packages("essurvey")
```

```{r, echo = FALSE, eval = TRUE}
library(essurvey)
```

Downloading the ESS data requires validating your email everytime you download data. We can set our email as an environment variable with `set_email`.

```{r}
set_email("your@email.com")
```

Once that's executed you can delete the previous line and any `import_*` call will look for the email automatically, stored as an environment variable.

Let's suppose you don't know which countries or rounds are available for the ESS. Then the `show_*` family of functions is your friend.

To find out which countries have participated you can use `show_countries()`

```{r, eval = TRUE}
show_countries()
```

This function actually looks up the countries in the ESS website. If new countries enter, this will automatically grab those countries as well. Let's check out Turkey. How many rounds has Turkey participated in? We can use `show_country_rounds()`

```{r, eval = TRUE}
tk_rnds <- show_country_rounds("Turkey")
tk_rnds
```
Note that country names are case sensitive. Use the exact name printed out by `show_countries()`

Using this information, we can download those specific rounds easily with `import_country`. Since `essurvey 1.0.0` all `ess_*` functions have been deprecated in favour of the `import_*` and `download_*` functions.

```{r}
turkey <-
  import_country(
    country = "Turkey",
    rounds = c(2, 4)
    )
```

`turkey` will now be a list of `length(rounds)` containing a data frame for each round. If you only specified one round, then all `import_*` functions return a data frame. `import_country` is useful for when you want to download specific rounds, but not all. To download all rounds for a country automatically you can use `import_all_cntrounds`.

```{r}
import_all_cntrounds("Turkey")
```

The `import_*` family is  concerned with downloading the data and thus always returns a list containing data frames unless only one round is specified, in which it returns a `tibble`. Conversely, the `show_*` family grabs information from the ESS website and always returns vectors.

## Download complete rounds

Similarly, we can use other functions to download rounds. To see which rounds are currently available, use `show_rounds`.

```{r, eval = TRUE}
show_rounds()
```

Similar to `show_countries`, `show_rounds` interactively looks up rounds in the ESS website, so any future rounds will automatically be included.

To download all available rounds, use `import_all_rounds`

```{r}
all_rounds <- import_all_rounds()
```

Alternatively, use `import_rounds` for selected ones.

```{r}
selected_rounds <- import_rounds(c(1, 3, 6))
```

## Downloading data for Stata, SPSS and SAS users

All `import_*` functions have an equivalent `download_*` function that allows the user to save the datasets in a specified folder in `'stata'`, `'spss'` or `'sas'` formats.

For example, to save round two from Turkey in a folder called `./my_folder`, we use:

```{r}
download_country("Turkey", 2,
                 output_dir = "./myfolder/")
```

By default it saves the data as `'stata'` files. Alternatively you can use `'spss'` or `'sas'`.

```{r}
download_country("Turkey", 2,
                 output_dir = "./myfolder/",
                 format = 'sas')
```

This will save the data to `./myfolder/ESS_Turkey` and inside that folder there will be the `ESS2` folder that contains the data.

## Correcting for missing values

Whenever you download the ESS data, it comes together with a script that recodes the values 6 = 'Not applicable', 7 = 'Refusal', 8 = 'Don't know', 9 = 'No answer' and 9 = 'Not available' as missings. However, that is the case for variables that have a scaling of 1-5. For variables which have a scaling from 1-10 the corresponding missings are 66, 77, and so on. At first glance new users might not know this and start calculating statistics with these variables such as...

```{r}
sp <- import_country("Spain", 1)
mean(sp$tvtot)
# 4.622406
```

..but that vector contains numbers such as `66`, `77`, that shouldn't be there. `recode_missings()` removes the corresponding missings for numeric variables as well as for character variables. It accepts the complete `tibble` and recodes all variables that should be recoded.

```{r}
new_coding <- recode_missings(sp)
mean(new_coding$tvtot, na.rm = TRUE)
# 4.527504
```

It also gives you the option of recoding only specific categories. For example...

```{r}
other_newcoding <- recode_missings(sp, c("Don't know", "Refusal"))
table(other_newcoding$tvpol)
#  0   1   2   3   4   5   6   7  66 
# 167 460 610 252  95  36  26  31  45 
```

...still has missing values but recoded the ones that were specified. I strongly suggest the user not to recode these categories as missing without looking at the data as there might be substantial diferences between people who didn't and who did answer questions. If the user is decided to do so, use `recode_missings` to recode everything and the corresponding `recode_*_missings` functions for numeric and character recodings separately. See the documentation of `?recode_missings` for more information.

## Analyzing ESS data

Be aware that for analyzing data from the ESS survey you should take into consideration the sampling and weights of the survey. The [survey](http://r-survey.r-forge.r-project.org/survey/) package provides very good support for this. A useful example comes from the work of Anthony Damico and Daniel Oberski [here](http://asdfree.com/european-social-survey-ess.html).

